---
title: "Evolutionary Dynamics Homework 9"
date:  "`r format(Sys.time(), '%B %d, %Y %H:%m')` (`r Sys.timezone()`)"
author: "Minghang Li"
mainfont: "KpRoman"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies:  [ "amsmath", "amsthm", "amssymb", "cancel", "unicode-math", "csquotes" ]
---

# Problem 1: Eden model dynamics

> Consider a two-dimensional Eden growth model with von Neumann neighborhood

## (b) Simulate different update rules

> Consider the R code at [https://git.io/vF0KH](https://git.io/vF0KH). What is the update implemented in the code? Modify the code to simulate the other two update rules from question (a) and then explain the difference between the spatial boundareis of the pouplation obtained for the different evolutionary dynamics.

```{r function_definition, include=FALSE}
################################### 
####### function definitions:
################################### 

# get the neighbourhood of a site:
nhood <- function(x, y, type = "von Neumann") {
  if(type != "Moore" && type != "von Neumann") stop("Neighbourhood type must be either Moore or von Neumann")
  if(type == "Moore") return(list(c(x - 1, y), c(x + 1, y), c(x, y - 1), c(x, y + 1),
                                                   c(x - 1, y - 1), c(x + 1, y + 1), c(x - 1, y + 1), c(x + 1, y - 1)))
  else return(list(c(x - 1, y), c(x + 1, y), c(x, y - 1), c(x, y + 1)))
}

# count the empty spaces in a neighbourhood:
count_spaces <- function(candidate, nhood_type) {
  neighbours <- nhood(candidate[1], candidate[2], nhood_type)
  count <- 0
  return(nhood_size - sum(sign(sapply(neighbours, function(e) sites[e[1], e[2]]))))
}
```

```{r parametrisation, include=FALSE}
################################### 
####### parametrisation:
###################################

grid_width <- 100 # width of the grid
nhood_type <- "von Neumann" # neighbourhood type
set.seed(5) # seed for random number generator (for replicable results)
occupied <- list(c(ceiling(grid_width/2), ceiling(grid_width/2))) # list of initially occupied sites
max_iter <- 1E4 # set a max number of iterations (just in case)
mutation_rate <- 0 # mutation rate per cell division
mutation_effect <- 0 # fitness effect per mutation
```

### Update rule implemented in the code

The update rule code is implemented in the following block:

```{r, eval=FALSE}
candidate <- sample(1:num_has_space,
                    1,
                    prob = unlist(how_many_spaces) * sapply(has_space, function(e)
                      sites[e[1], e[2]]))
```

It basically selects a cell that will divide in this iteration, with a probability
that depends on both the number of empty spaces in the neighbourhood and the fitness
of the cell itself. This implementation is most similar to the _bond focused_ update
rule in question (a), where we randomly choose an $S_1$ (occupied) site with probability
proportional to the number of adjoining $S_0$ (empty) sites.


### Simulate the other two update rules

#### Available site-focused

- _Available site-focused_ update rule: randomly choose an $S_0$ that adjoins at least one $S_1$ site, and switch it from $S_0$ to $S_1$.

```{r available_site_focused}
###################################
####### initialization:
###################################

# Initialize site states:
sites <- matrix(0, nrow = grid_width, ncol = grid_width)
num_occupied <- length(occupied)
for (i in 1:num_occupied)
  sites[occupied[[i]][1], occupied[[i]][2]] <- 1

# neighbourhood size:
nhood_size <- length(nhood(1, 1, nhood_type))

# Initialize list of empty sites adjacent to occupied ones:
num_empty_adj <- 0
empty_adj <- list()
index_map <- matrix(NA, nrow = grid_width, ncol = grid_width)

# Find initial empty sites adjacent to occupied ones:
for (x in 2:(grid_width - 1))
  for (y in 2:(grid_width - 1)) {
    if (sites[x, y] < 1) {
      # if site is empty
      neighbours <- nhood(x, y, nhood_type)
      if (any(sapply(neighbours, function(n)
        sites[n[1], n[2]] > 0))) {
        # has occupied neighbor
        num_empty_adj <- num_empty_adj + 1
        empty_adj[[num_empty_adj]] <- c(x, y)
        index_map[x, y] <- num_empty_adj
      }
    }
  }

# Initialize timer and output:
timer <- 0
output_df <- data.frame(Time = timer, Population = num_occupied)

###################################
####### run the model:
###################################

for (iter in 1:max_iter) {
  if (num_empty_adj == 0)
    break
  
  # Pick random empty site adjacent to occupied:
  candidate_idx <- sample(num_empty_adj, 1)
  new_site <- empty_adj[[candidate_idx]]
  
  # Switch from empty to occupied:
  sites[new_site[1], new_site[2]] <- 1
  num_occupied <- num_occupied + 1
  
  # Remove chosen site from empty_adj list:
  empty_adj[[candidate_idx]] <- empty_adj[[num_empty_adj]]
  index_map[empty_adj[[num_empty_adj]][1], empty_adj[[num_empty_adj]][2]] <- candidate_idx
  empty_adj[[num_empty_adj]] <- NULL
  index_map[new_site[1], new_site[2]] <- NA
  num_empty_adj <- num_empty_adj - 1
  
  # Add new empty neighbors to empty_adj list:
  neighbours <- nhood(new_site[1], new_site[2], nhood_type)
  for (n in neighbours) {
    if (n[1] >= 1 &&
        n[1] <= grid_width && n[2] >= 1 && n[2] <= grid_width &&
        sites[n[1], n[2]] < 1 && is.na(index_map[n[1], n[2]])) {
      num_empty_adj <- num_empty_adj + 1
      empty_adj[[num_empty_adj]] <- n
      index_map[n[1], n[2]] <- num_empty_adj
    }
  }
  
  # Update time (using simpler rate based on population):
  timer <- timer + rexp(1, num_occupied)
  
  # Record time and population:
  output_df <- rbind(output_df, c(timer, num_occupied))
  
  # Check boundary:
  if (new_site[1] >= grid_width || new_site[1] <= 1 ||
      new_site[2] >= grid_width || new_site[2] <= 1)
    break
}
```

```{r plottt, echo=FALSE, fig.width=8, fig.height=8}
################################### 
####### plot the results:
################################### 

# par(mar=c(2,2,2,2)) 
par(mfrow = c(2, 2))

# plot the occupied sites,
# after changing the value assigned to empty sites (for clearer plotting):
num_types <- length(unique(as.numeric(sites))) # number of unique fitness values
sites_for_plot <- sites
if(num_types > 2) sites_for_plot[which(sites_for_plot == 0)] <- min(sites[which(sites > 0)]) - (max(sites) - min(sites[which(sites > 0)])) / 3
image(sites_for_plot, main = "Max relative\nfitness", col = c("black", heat.colors(num_types - 1)))

# initial list of occupied sites that are next to at least one empty site,
# and initial list of how many empty neighbours each has:
num_has_space <- 0
has_space <- list()
how_many_spaces <- list()
for(x in 2:(grid_width - 1)) for(y in 2:(grid_width - 1)) {
  candidate <- c(x, y)
  spaces <- count_spaces(candidate, nhood_type)
  # if not all neighbours are occupied then add the site to the list:
  if(sites[candidate[1], candidate[2]] > 0 && spaces > 0) {
    num_has_space <- num_has_space + 1 # number of sites that have space
    has_space[[num_has_space]] <- candidate # list of sites that have space
    how_many_spaces[[num_has_space]] <- spaces # list of numbers of spaces for the sites that have space
    index_map[candidate[1], candidate[2]] <- num_has_space # map from from grid sites to list entries
  }
}

# plot the numbers of empty neighbours:
space_grid <- matrix(0, nrow = grid_width, ncol = grid_width)
for(x in 1:grid_width) for(y in 1:grid_width) {
  index <- index_map[x, y]
  if(!is.na(index)) space_grid[x, y] <- how_many_spaces[[index]]
}
image(space_grid, main = "Number of\nempty neighbours", col = c("black", heat.colors(nhood_size)))

# plot the growth curves of population and equivalent radius:
plot(Population ~ Time, data = output_df, type = "l", main = "Population growth")
plot(sqrt(Population / pi) ~ Time, data = output_df, type = "l", ylab = "Equivalent radius", main = "Equivalent\nradius growth")
```

#### Cell-focused

- _Cell-focused_ update rule: randomly choose an $S_1$ site that adjoins at least one $S_0$ site, then sample uniformly an $S_0$ neighbour and switch it to $S_1$.

Basically just remove the scaling using `how_many_spaces` in the `sample` function.

```{r eval=FALSE}
candidate <- sample(1:num_has_space, 1, prob = sapply(has_space, function(e)
  sites[e[1], e[2]]))
```


```{r run_the_model, include=FALSE}
################################### 
####### intialisation:
###################################

# initialise the site states:
sites <- matrix(0, nrow = grid_width, ncol = grid_width)
num_occupied <- length(occupied) # number of occupied sites
for(i in 1:num_occupied) sites[occupied[[i]][1], occupied[[i]][2]] <- 1

# neighbourhood size:
nhood_size <- length(nhood(1, 1, nhood_type))

# initialise the array that maps from grid sites to list entries:
index_map <- matrix(NA, nrow = grid_width, ncol = grid_width)

# initial list of occupied sites that are next to at least one empty site,
# and initial list of how many empty neighbours each has:
num_has_space <- 0
has_space <- list()
how_many_spaces <- list()
for(x in 2:(grid_width - 1)) for(y in 2:(grid_width - 1)) {
  candidate <- c(x, y)
  spaces <- count_spaces(candidate, nhood_type)
  # if not all neighbours are occupied then add the site to the list:
  if(sites[candidate[1], candidate[2]] > 0 && spaces > 0) {
    num_has_space <- num_has_space + 1 # number of sites that have space
    has_space[[num_has_space]] <- candidate # list of sites that have space
    how_many_spaces[[num_has_space]] <- spaces # list of numbers of spaces for the sites that have space
    index_map[candidate[1], candidate[2]] <- num_has_space # map from from grid sites to list entries
  }
}

# initialise the timer and the output dataframe:
timer <- 0
output_df <- data.frame(Time = timer, Population = num_occupied)
################################### 
####### run the model:
################################### 
for(iter in 1:max_iter) {
  # pick a cell to divide:
  candidate <- sample(1:num_has_space, 1, prob = sapply(has_space, function(e) sites[e[1], e[2]]))
  # find the cell's neighbours:
  neighbours <- nhood(has_space[[candidate]][1], has_space[[candidate]][2], nhood_type)
  # randomise the order in which neighbours will be tested:
  test_order <- sample(1:length(neighbours))
  
  # test neighbours one by one:
  num_tested <- 0
  for(i in test_order) {
    # keep count of the number of occupied neighbours (for error checking):
    if(sites[neighbours[[i]][1], neighbours[[i]][2]] > 0) num_tested <- num_tested + 1
    # if the neighbouring site is empty then make it occupied:
    else {
      num_occupied <- num_occupied + 1
      new_site <- neighbours[[i]]
      parent_site <- has_space[[candidate]]
      # mutation:
      if(runif(1) < mutation_rate) sites[new_site[1], new_site[2]] <- (1 + mutation_effect) * sites[parent_site[1], parent_site[2]]
      else sites[new_site[1], new_site[2]] <- sites[parent_site[1], parent_site[2]]
      break
    }
  }
  # error check:
  if(num_tested == length(neighbours)) stop("Site wrongly listed as having space to divide")
  
  # time between events is drawn from an exponential distribution,
  # consistent with the Gillespie algorithm:
  sum_of_rates <- sum(sapply(has_space, function(e) sites[e[1], e[2]]) * unlist(how_many_spaces) / nhood_size)
  timer <- timer + rexp(1, sum_of_rates)
  
  # record the time and the population size:
  output_df <- rbind(output_df, c(timer, num_occupied))
  
  # stop if a cell has reached the edge of the grid:
  if(new_site[1] >= grid_width || new_site[1] <= 1 || new_site[2] >= grid_width || new_site[2] <= 1) break
  
  # update the other lists for the newly occupied site, if it has empty neighbours:
  spaces <- count_spaces(new_site, nhood_type)
  if(spaces > 0) {
    num_has_space <- num_has_space + 1 # number of sites that have space
    has_space[[num_has_space]] <- new_site # list of sites that have space
    how_many_spaces[[num_has_space]] <- spaces # list of numbers of spaces for the sites that have space
    index_map[new_site[1], new_site[2]] <- num_has_space # map from from grid sites to list entries
  }
  
  # update list entries for neighbouring occupied sites:
  neighbours <- nhood(new_site[1], new_site[2], nhood_type)
  for(candidate in neighbours) if(sites[candidate[1], candidate[2]] > 0) {
    index <- index_map[candidate[1], candidate[2]]
    # error check:
    if(is.na(index)) stop("Site wrongly listed in index_map")
    
    # decrement the number of empty spaces neighbouring the site:
    how_many_spaces[[index]] <- how_many_spaces[[index]] - 1
    
    # if there are no neighbouring empty sites then remove the site from the lists:
    if(how_many_spaces[[index]] == 0) {
      has_space[[index]] <- NULL # removes the entry and shifts subsequent entries down one slot
      how_many_spaces[[index]] <- NULL # removes the entry and shifts subsequent entries down one slot
      num_has_space <- num_has_space - 1
      index_map[candidate[1], candidate[2]] <- NA
      index_map[which(index_map > index)] <- index_map[which(index_map > index)] - 1 # because entries have shifted by one slot
    }
  }
}
```

```{r plot_the_results, echo=FALSE, fig.width=8, fig.height=8}
################################### 
####### plot the results:
################################### 

# par(mar=c(2,2,2,2)) 
par(mfrow = c(2, 2))

# plot the occupied sites,
# after changing the value assigned to empty sites (for clearer plotting):
num_types <- length(unique(as.numeric(sites))) # number of unique fitness values
sites_for_plot <- sites
if(num_types > 2) sites_for_plot[which(sites_for_plot == 0)] <- min(sites[which(sites > 0)]) - (max(sites) - min(sites[which(sites > 0)])) / 3
image(sites_for_plot, main = "Max relative\nfitness", col = c("black", heat.colors(num_types - 1)))

# plot the numbers of empty neighbours:
space_grid <- matrix(0, nrow = grid_width, ncol = grid_width)
for(x in 1:grid_width) for(y in 1:grid_width) {
  index <- index_map[x, y]
  if(!is.na(index)) space_grid[x, y] <- how_many_spaces[[index]]
}
image(space_grid, main = "Number of\nempty neighbours", col = c("black", heat.colors(nhood_size)))

# plot the growth curves of population and equivalent radius:
plot(Population ~ Time, data = output_df, type = "l", main = "Population growth")
plot(sqrt(Population / pi) ~ Time, data = output_df, type = "l", ylab = "Equivalent radius", main = "Equivalent\nradius growth")
```

### Analysis

Available site-focused update rule grows the fastest to grid edge, but creates spiky, irregular boundaries and shows holes within occupied areas. It suggests that random selection of empty sites leads to uneven growth. Cell-focused update rule has rather slow growth rate, creates smooth, compact patterns.


# Problem 2: Diffusion approximation of a spatial Moran model

> Consider the spatial Moran model for a mutation spreading through an infinite row of demes. Let $\mu$ denote the death rate, $s$ the fitness advantage of the mutant, $m$ the dispersion probability, and $N$ the number of individuals per deme. Assume initially that $n_i = N, \forall i \leq 0$ and $n_i = 0$ otherwise, where $n_i$ is the number of mutants in deme i. The rate at which mutants in a deme die is $\mu n_i$. Once an individual has died, it is replaced by the offspring of a wild type individual of its own deme with probability

$$
P_1 = \frac{N-n_i}{(1+s)n_i + (N - n_i)} \approx \frac{N-n_i}{N},
$$

> and it is replaced by the offspring of a wild type individual from any of the neighbouring demes with probability

$$
P_2 = \frac{m(N - n_{i-1})}{2N} + \frac{m(N - n_{i+1})}{2N}
\approx
\frac{m(2N - (n_{i-1} + n_{i+1}))}{2N}.
$$

## (a) Using these results, show that the rate for the number of mutants in deme $i$ decreasing by one individual is as follows:

$$
W_i^{-}(\mathbf{n}) = \frac{\mu}{N} n_i  \left(
  (N - n_i) - \frac{m}{2} n_i''
\right)
$$

where $n_i'' = n_{i-1} + n_{i+1} - 2n_i$.

---

\begin{proof}
$W_i^{-}(\mathbf{n})$ is basically $\mu n_i ((1-m)P_1 + P_2)$.
$$
\begin{aligned}
W_i^{-}(\mathbf{n}) &= \mu n_i \left( 
(1-m) \cdot \frac{N-n_i}{N} + 
\frac{m(2N - (n_{i-1} + n_{i+1}))}{2N} \right) \\
&= \frac{\mu n_i}{N}\left(
  N - n_i - m(N - n_i) + \frac{m(2N - (n_{i-1} + n_{i+1}))}{2}
\right) \\
&= \frac{\mu n_i}{N}\left(
  N - n_i - m\frac{
    \cancel{2N} - 2n_i - \cancel{2N} + n_{i-1} + n_{i+1}
  }{2}
\right) \\
&= \frac{\mu n_i}{N}\left(
  N - n_i - \frac{m}{2}(n_{i-1} + n_{i+1} - 2n_i)
\right) \\\
&= \frac{\mu}{N} n_i  \left(
  (N - n_i) - \frac{m}{2} n_i''
\right)
\end{aligned}
$$
\end{proof}

## (b) Show that the rate of expectation of the number of mutants in deme $i$ is as follows:

> In general,

$$
\frac{d \mathbb{E}[n_i]}{dt} 
  = \mathbb{E}\left[W_i^{+}(\mathbf{n}) - W_i^{-}(\mathbf{n}) \right]
$$

> where $\mathbb{E}[X]$ denotes the expected value of a random variable $X$. Using the expression of $W_i^{+}(\mathbf{n})$ given in the lecture and applying the approximation $\mathbb{E}[n_in_k] \approx \mathbb{E}[n_i]\mathbb{E}[n_k]$ (called the _mean-field approximation_), show that

$$
\frac{d \mathbb{E}[n_i]}{dt}
  = \frac{\mu m}{2} \mathbb{E}[n_i''] + 
    \frac{s\mu}{N} (N - \mathbb{E}[n_i]) \left(
      \mathbb{E}[n_i] + \frac{m}{2} \mathbb{E}[n_i'']
    \right)
$$

---

\begin{proof}
From lecture, we know that 
$$
W_i^{+}(\mathbf{n}) = 
  \frac{\mu(1+s)}{N} (N-n_i) \left[
  n_i + \frac{m}{2} n_i''\right
].
$$

Then we have 

$$
\begin{aligned}
\frac{d \mathbb{E}[n_i]}{dt}
  &= \mathbb{E}\left[W_i^{+}(\mathbf{n}) - W_i^{-}(\mathbf{n}) \right] \\
  &= \mathbb{E}\left[
  \frac{\mu(1+s)}{N} (N-n_i) \left[
  n_i + \frac{m}{2} n_i''\right] - 
  \frac{\mu}{N} n_i \left(
    (N - n_i) - \frac{m}{2} n_i''
    \right)
  \right] \\
  &(\text{Denote } \frac{m}{2}n_i'' \text{ as } M \text{ for succinctness}) \\
  &= \mathbb{E}\left[
  \frac{\mu(1+s)}{N} (N-n_i) \left[n_i + M\right] - \frac{\mu}{N}n_i(N - n_i - M)
  \right] \\
  &= \mathbb{E}\left[
    \frac{s\mu}{N}(N - n_i)(n_i + M)
  \right] + \mathbb{E} \left[
    \frac{\mu}{N} \left[
      (N - n_i)(n_i + M) - n_i(N - n_i - M)
    \right]
  \right] \\
  &= \mathbb{E}\left[
    \frac{s\mu}{N}(N - n_i)(n_i + M)
  \right] + \mathbb{E} \left[
    \frac{\mu}{N} \left[
      (\cancel{Nn_i} + NM -\cancel{n_i^2} - \cancel{n_iM} - \cancel{n_iN} + \cancel{n_i^2} + \cancel{n_iM})
    \right]
  \right] \\
  &(\text{Change } M \text{ back to } \frac{m}{2}n_i'') \\
  &= \frac{s\mu}{N} \mathbb{E}\left[
    (N - n_i)\left(n_i + \frac{m}{2}n_i''\right)
    \right] + \mu \mathbb{E} \left[
      \frac{m}{2}n_i''
    \right] \\
  &= \frac{s\mu}{N} (N - \mathbb{E}[n_i]) \left(
    \mathbb{E}[n_i] + \frac{m}{2} \mathbb{E}[n_i'']
  \right) + \frac{\mu m}{2} \mathbb{E}[n_i'']
\end{aligned}
$$
\end{proof}

## (c) Show that the process can be approximate by diffusion equation.

> We can approximate distance along the row of demes using the continuous variable $x = li$, where $l$ is the deme width. Setting $u(x) = \mathbb{E}[n_i]/N$, use the previous approximation to show that process proved in (b) can  be approximated by the diffusion equation

$$
\frac{\partial u}{\partial t} =
  D[1+s(1-u)] \frac{\partial^2 u}{\partial x^2} + \mu s u (1-u)
$$

> where diffusion coefficient $D = uml^2 /2$.
>
> _Note_: we assume that $\partial u(x) / \partial i$ is the expected difference in $u$ between the $i$th and $(i+i)$th deme and that $\partial^2 u(x) / \partial i^2$ is the expected difference in $\partial u(x) / \partial i$ between the $i$th and $(i+i)$th deme. Compute $\partial u(x) / \partial i^2$ and perform a change of variables.

---

\begin{proof}
$$
\begin{aligned}
\frac{d \mathbb{E}[n_i]}{dt}
&= \frac{s\mu}{N} (N - \mathbb{E}[n_i]) \left(
    \mathbb{E}[n_i] + \frac{m}{2} \mathbb{E}[n_i'']
  \right) + \frac{\mu m}{2} \mathbb{E}[n_i''] \\
\frac{1}{N} \frac{d \mathbb{E}[n_i]}{dt} &= 
  s \mu \left(1 - \frac{\mathbb{E}[n_i]}{N}\right)
  \left(\frac{\mathbb{E}[n_i]}{N} + \frac{m}{2}\frac{\mathbb{E}[n_i'']}{N}\right) +
  \frac{\mu m}{2} \frac{\mathbb{E}[n_i'']}{N} \\
\frac{\partial u}{\partial t} &= s \mu (1 - u) \left(u + \frac{m}{2} \frac{\mathbb{E}[n_i'']}{N}\right) + \frac{\mu m}{2} \frac{\mathbb{E}[n_i'']}{N} \\
\frac{\partial u}{\partial t} &= \mu s u (1 - u) + \frac{\mu m}{2}\frac{\mathbb{E}[n_i'']}{N} [1 + s(1-u)]
\end{aligned}
$$

Now we just need to tidy $\mathbb{E}[n_i'']/N$ up!

$$
\begin{aligned}
\frac{\mathbb{E[n_i'']}}{N}
  &= \frac{\mathbb{E}[n_{i+1} - n_i]}{N} - \frac{\mathbb{E}[n_{i} - n_{i-1}]}{N} \\
  &= \mathbb{E}[u_{i+1} - u_i] - \mathbb{E}[u_i - u_{i-1}] \\
  &= \frac{\partial u}{\partial i} - \frac{\partial u}{\partial i-1} \\
  &= \frac{\partial^2 u}{\partial i^2}
\end{aligned}
$$
Change of variables:

$$
\frac{\partial^2 u}{\partial x^2} 
  = \frac{\partial^2 u}{\partial i^2} \frac{\partial^2 i}{\partial x^2}
  = \frac{\partial^2 u}{\partial i^2} \frac{1}{l^2}
$$

Combine everthing we have:

$$
\begin{aligned}
\frac{\partial u}{\partial t} 
&= \mu s u (1 - u) + \frac{\mu m}{2}\frac{\mathbb{E}[n_i'']}{N} [1 + s(1-u)]  \\
&= \mu s u (1 - u) + \frac{\mu m l^2}{2}\frac{\partial^2u}{\partial x^2} [1 + s(1-u)]
\end{aligned}
$$
\end{proof}

## (d) See below.

```{r}
# Parameters
dx <- 0.1  # Space step
dt <- 0.001  # Time step
L <- 16  # Domain length [-L, L]
T <- 5  # Total time
mu <- 0.1  # death rate
m <- 0.3 # Dispersion probability
s <- 0  # Fitness advantage
```

```{r}
simulate <- function(
  dx=0.1, dt=0.001, L=2, T=1, mu=1, m=0.2, s=0  
) {
  x <- seq(-L, L, by=dx)
  t <- seq(0, T, by=dt)
  nx <- length(x)
  nt <- length(t)
  D <- mu*m*L^2/2  # Diffusion coefficient
  
  u <- matrix(0, nrow=nx, ncol=nt)
  u[x <= 0, 1] <- 1  # Set u=1 (meaning n=N) for x≤0
  u[x > 0, 1] <- 0   # Set u=0 (meaning n=0) for x>0
  
  for(j in 1:(nt-1)) {
    for(i in 2:(nx-1)) {
      d2u_dx2 <- (u[i+1,j] - 2*u[i,j] + u[i-1,j])/dx^2
      
      u[i,j+1] <- u[i,j] + dt * (
        D * (1 + s*(1-u[i,j])) * d2u_dx2 + 
        mu * u[i,j] * (1-u[i,j])
      )
      
      u[i,j+1] <- max(0, min(1, u[i,j+1]))
    }
    
    # Boundary conditions (Dirichlet)
    u[1,j+1] <- u[1,j]
    u[nx,j+1] <- u[nx,j]
  }
  return(
    list(
      x=x, t=t, u=u
    )
  )
}
```

```{r}
plot_diffusion <- function(x, t, u, times_to_plot = c(0, 0.2, 0.5, 1), s=0) {
  # Plot results at different time points
  
  colors <- rainbow(length(times_to_plot))
  plot(
    x,
    u[, 1],
    type = 'l',
    ylim = c(0, 1),
    col = colors[1],
    xlab = 'Position',
    ylab = 'u',
    main = paste('Diffusion-Growth, s =', s)
  )
  
  for (i in 2:length(times_to_plot)) {
    j <- which.min(abs(t - times_to_plot[i]))
    lines(x, u[, j], col = colors[i])
  }
  
  legend(
    'topright',
    legend = paste('t =', times_to_plot),
    col = colors,
    lty = 1,
    cex = 0.8
  )
}
```

```{r}
res <- simulate(dx, dt, L, T, mu, m, s)
plot_diffusion(res$x, res$t, res$u, seq(0, T), s=0)
```

```{r}
res <- simulate(dx, dt, L, T, mu, m, s=0.5)
plot_diffusion(res$x, res$t, res$u, seq(0, T), s=0.5)
```

In the case $s = 0$ the shape of the function will remain unchanged, while for $s > 0$, there will be a shift towards the right, like a wave. The biology intuition is that the mutant with more fitness will expand over neighbouring demes with a speed proportional to its selective advantage and the diffusion constant.

## (e) See below.

$D = μml^2/2$ is the diffusion coefficient, where:

$m$ is the dispersal probability that:

- Controls how easily individuals move between demes
- Higher $m$ means faster spatial spread

$μ$ is the death rate:

- Affects both diffusion speed and population growth
- Appears in the growth term $\mu su(1-u)$

$s$ as the selective advantage:

- Measures fitness advantage of mutants
- When $s = 0$, only random diffusion occurs
- When $s > 0$, adds directional movement to the wave
