---
title: "Evolutionary Dynamics Exercise 1"
date:  "`r format(Sys.time(), '%B %d, %Y')` (`r Sys.timezone()`)"
author: "Minghang Li"
mainfont: "KpRoman"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies:  [ "amsmath", "amsthm", "amssymb", "cancel", "unicode-math" ]
---

This solution PDF is written in R markdown.

```{r session_info}
sessionInfo()
```

```{r include}
library(foreach)
library(ggplot2)
```

# Problem 1: Discrete time

Suppose you have a difference equation $x_{t+1} = f(x_t)$ of a discrete time model with

$$
f(x) = 5 x^2 (1-x).
$$

## (a) Determine the equilibrium points $x^*$ of the system.

At equilibrium points we have $x_{t+1} = x_t = x^*$, which means

$$
\begin{aligned}
x^* - 5 {x^*}^2 (1- x^*) &= 0\\
5{x^*}^3 - 5 {x^*}^2 + x^* &= 0 \\
x^* (5{x^*}^2 - 5x^* + 1) &= 0 \\
\end{aligned}
$$

Solving the equation gives solutions:

$$
x_1^* = 0, \qquad
x_2^* = \frac{1}{2} - \frac{1}{2\sqrt{5}} \qquad
x_3^* = \frac{1}{2} + \frac{1}{2\sqrt{5}}
$$

## (b) Which of the equilibrium points $x^*$ are stable?

For an equilibrium to be stable we need to have the absolute value of the derivative of $|f'(x^*)| < 1$.

$$
f(x) = 5 x^2 (1-x) = -5 x^3 + 5x^2 \\
f'(x) = -15 x^2 + 10 x
$$

Plug in $x_1^* = 0$, $x_2^* = \frac{1}{2} - \frac{1}{2\sqrt{5}}$, $x_3^* = \frac{1}{2} + \frac{1}{2\sqrt{5}}$, we have

$$
\begin{aligned}
\left| f'(x_1^* )\right| &= 0 < 1 & \text{(stable)} \\
\left| f'(x_2^* )\right| &= \frac{1}{2} + \frac{\sqrt{5}}{2} > 1 & \text{(unstable)} \\
\left| f'(x_3^* )\right| &= \frac{\sqrt{5}}{2} - \frac{1}{2} < 1 & \text{(stable)}
\end{aligned}
$$


# Problem 2: Continuous time

Consider the case:

$$
\frac{dx(t)}{dt} = f(x) = 3x(x-1)(x-2).
$$

## (a) Determine the equilibrium points $x^*$ of the system.

At the equilibrium, we have

$$
\frac{dx^*(t)}{dt} = f(x^*) = 3x^*(x^*-1)(x^*-2) = 0
$$

which yields:

$$
x_1^* = 0, \qquad x_2^* = 1, \qquad x_3^* = 2
$$

## (b) Which of the equiliubrium points $x^*$ are stable?

$$
f'(x^*) = 3(x^* -1)(x^*-2) + 3x^*(x^* -2) + 3x^*(x^* - 1)
$$

Plug in $x_1^* = 0$, $x_2^* = 1$, $x_3^* = 2$, we have

$$
\begin{aligned}
f'(x_1^*) &= 6  > 0& (\text{unstable}) \\
f'(x_2^*) &= -3 < 0 & (\text{stable}) \\
f'(x_3^*) &= 6  < 0& (\text{unstable})
\end{aligned}
$$

# Problem 3: Logistic difference equation

In a discrete time model for population growth, the value $x$ (number of cells divided by the maximum number supported by the habitat) at time $t + 1$ is calculated from the value at time $t$ according to the difference equation:

$$
x_{t+1} = rx_t(1-x_t)
$$

## (a) Determine the equilibrium points $x^*$ of the system.

At equilibrium we have $x_{t+1} = x_t = x^*$, which means:

$$
\begin{aligned}
x^* &= r x^* (1-x^*) \\
x^*(rx^* - (r - 1)) &= 0
\end{aligned}
$$

Solving the equation gives:

$$
x_1^* = 0, \qquad x_2^* = \frac{r-1}{r}
$$

## (b) Are the equilibrium points table for $r = 0.9$, $r = 1.9$, $r = 2.9$?

As $f(x) = rx(1-x)$, we have:

$$
f'(x) = r - 2rx
$$

Plug in $x_1^* = 0$, $x_2^* = \frac{r - 1}{r}$ with different $r$ values into the equation, we have:

### (1) $r = 0.9$

With $r = 0.9$, $x_2^* = \frac{0.9 - 1}{0.9} = \frac{-1}{9}$.

$$
\begin{aligned}
\left| f'(x_1^* )\right| &= \left| f'(0)\right| = 0.9 < 1 & \text{(stable)} \\
\left| f'(x_2^* )\right| &= \left| f'\left(\frac{1}{9}\right)
\right| = \left|
\frac{9}{10} + 2 \cdot \frac{9}{10} \cdot \frac{1}{9}
\right| = \frac{11}{10} > 1 & (\text{unstable})
\end{aligned}
$$

### (2) $r = 1.9$

With $r = 1.9$, $x_2^* = \frac{1.9 - 1}{1.9} = \frac{9}{19}$.

$$
\begin{aligned}
\left| f'(x_1^*) \right| &= \left| f'(0)\right| = 1.9 > 1 & \text{(unstable)} \\
\left| f'(x_2^*) \right| &= \left|
  f'\left(\frac{9}{19}\right)
\right| = \left|
\frac{19}{10} - 2 \cdot \frac{19}{10} \cdot \frac{9}{19}
\right| = \frac{1}{10} < 1 & (\text{stable})
\end{aligned}
$$

### (3) $r = 2.9$

With $r = 2.9$, $x_2^* = \frac{2.9 - 1}{2.9} = \frac{19}{29}$.

$$
\begin{aligned}
\left| f'(x_1^*) \right| &= \left| f'(0)\right| = 2.9 > 1 & \text{(unstable)} \\
\left| f'(x_2^*)\right| &= \left|
  f'\left(\frac{19}{29}\right)
\right| = \left|
  \frac{29}{10} - 2 \cdot \frac{29}{10} \cdot \frac{19}{29}
\right| = \frac{9}{10} < 1 & (\text{stable})
\end{aligned}
$$

## (c) Confirm this by numerically iterating the difference equation

```{r numerical_simulation}
sim_diffeq <- function(ini, r, steps=100) {
  x <- rep(0, steps)
  x[1] <- ini
  foreach (i=2:steps) %do% {
    x[i] <- r * x[i - 1] * (1 - x[i - 1])
  }
  return(x)
}
```

```{r run_sim}
sim1 <- sim_diffeq(0.5, 0.9)
sim2 <- sim_diffeq(0.5, 1.9)
sim3 <- sim_diffeq(0.5, 2.9)
```

```{r plot}
par(mar = c(4, 4, 2, 0.5))
par(mgp = c(2.5, 1, 0))
# par(cex.lab = 1.25)
par(mfrow=c(1,3))
plot(sim1,
     xlab = "step",
     ylab = "x",
     main = "r=0.9",
     col = "dodgerblue",
     pch = 20,
     ylim = c(0,1)
)
abline(h = 0, col = "orange", lwd=2)

plot(sim2,
     xlab = "step",
     ylab = "x",
     main = "r=1.9",
     col = "dodgerblue",
     pch = 20,
     ylim = c(0,1)
)
abline(h = 9/19, col = "orange", lwd=2)

plot(sim3,
     xlab = "step",
     ylab = "x",
     main = "r=2.9",
     col = "dodgerblue",
     pch = 20,
     ylim = c(0,1)
)
abline(h = 19/29, col = "orange", lwd=2)
```

## (d) Examine the stability and behavior for $r = 3.2$.

The system approaches permanent oscillation between two values (can be computed using the same technique used in previous sections).

```{r run_sim_r32}
par(mar = c(4, 4, 2, 0.5))
par(mgp = c(2.5, 1, 0))
par(cex.lab = 1.25)
par(mfrow=c(1,2))

sim4 <- sim_diffeq(0.5, 3.2)

xstm1 <- sim4[-length(sim4)]
xst <- sim4[-1]

plot(sim4,
     xlab = "step",
     ylab = "x",
     main = "Simulation (r=3.2)",
     col = "dodgerblue",
     pch = 20,
     ylim = c(0,1)
)

# Poincare
plot(xstm1,
     xst,
     xlab = expression(x[t-1]),
     ylab = expression(x[t]),
     main = "Poincare plot (r=3.2)",
     col = "dodgerblue",
     pch = 20,
     xlim = c(0,1),
     ylim = c(0,1)
)
lines(xstm1, xst, col = "dodgerblue", lwd=0.5)
abline(b = 1, a = 0, col = "orange", lwd =2)
```

## (e) What happens for $r = 3.9$?

The system exhibits chaotic behavior.

```{r run_sim_r39}
par(mar = c(4, 3, 2, 0.5))
par(mgp = c(2.5, 1, 0))
par(cex.lab = 1.25)
par(mfrow=c(1,2))

sim4 <- sim_diffeq(0.5, 3.9)

xstm1 <- sim4[-length(sim4)]
xst <- sim4[-1]

plot(sim4,
     xlab = "step",
     ylab = "x",
     main = "Simulation (r=3.9)",
     col = "dodgerblue",
     pch = 20,
     ylim = c(0,1)
)

# Poincare
plot(xstm1,
     xst,
     xlab = expression(x[t-1]),
     ylab = expression(x[t]),
     main = "Poincare plot (r=3.9)",
     col = "dodgerblue",
     pch = 20,
     xlim = c(0,1),
     ylim = c(0,1)
)
lines(xstm1, xst, col = "dodgerblue", lwd=0.5)
abline(b = 1, a = 0, col = "orange", lwd =2)
```

## (r) Describe what happens for $r > 4$. Is the model still biologically plausible?

Beyond r = 4, almost all initial values eventually leave the interval [0,1] and diverge. It is not biologically plausible.

# Problem 4: Logistic growth in continuous time

The logistic model for population growth is:

$$
\frac{dx(t)}{dt} = \lambda x(t) \left(
  1 - \frac{x(t)}{K}
\right)
$$

## (a) Show the solution of logistic growth by direct integral

$$
x(t) = \frac{K x_0 e^{\lambda t}}{ K + x_0 (e^{\lambda t} - 1)}
$$
This is a very basic linear differential equation exercise.

$$
\begin{aligned}
\int\frac{dx(t)}{x(t)\left(1- \frac{x(t)}{K}\right)} &= \int \lambda dt \\
\int \frac{dx(t)}{x(t)} + \int \frac{dx(t)}{K - x(t)} &= \int \lambda dt \\
\ln |x(t)| - \ln |K - x(t)| &= \lambda t + C \\
\ln \left| \frac{K - x(t)}{x(t)} \right| &= -(\lambda t + C) \\
\left|
  \frac{K - x(t)}{x(t)}
\right| &= e^{-(\lambda t + C)} \\
\frac{K - x(t)}{x(t)} &= C_0 \cdot e^{-\lambda t}
\end{aligned}
$$
Rewrite the equation we get:

$$
x(t) = \frac{K}{1+C_0e^{-\lambda t}}
$$

To get the final form, we need to replace $C_0$ with some expression with $x_0$, which is the initial condition of the population at $t = 0$:

$$
\frac{K - x_0}{x_0} = C_0
$$

Plug it back to the equation, we have:

$$
\begin{aligned}
x(t) &= \frac{K}{ 1 + \frac{K - x_0}{x_0} e^{-\lambda t}} \\
&= \frac{K x_0 e^{\lambda t}}{ K + x_0 (e^{\lambda t} - 1)} \\
\end{aligned}
$$

## (b) Find the equilibrium points of the systems and discuss their stability

At equilibrium points we have $\frac{dx(t)}{dt} = f(x) = 0$, i.e.:

$$
f(x^*) = \lambda x^* \left(1 - \frac{x^*}{K}\right) = 0 \Longrightarrow x_1^* = 0, \quad x_2^* = K
$$

$$
f'(x) = \lambda \left(1- \frac{x}{K}\right) - \frac{\lambda x}{K} = \lambda \left( 1 - \frac{2x}{K} \right) 
$$

At equilibrium points, we have

$$
\begin{aligned}
f'(x_1^*) &= f'(0) = \lambda \\
f'(x_2^*) &= f'(K) = -\lambda
\end{aligned}
$$

Interestingly, we will then only have one of the two equilibrium points to be stable. If $\lambda > 0$, then $x_1^* = 1$ is unstable and $x_2^* = K$ is stable; while if $\lambda < 0$ we have the exact opposite.

## (c) Demostrate the results above for $K = 3$ and a series of time points.

```{r logistic_sim}
logistic_sim <- function (ini, lambda, K, steps=100) {
  x <- rep(0, steps)
  x[1] <- ini
  foreach (i=2:steps) %do% {
    x[i] <- lambda * x[i-1] * (1 - (x[i-1])/K) + x[i-1]
  }
  return(x)
}
```

```{r run_logistic}
par(mar = c(4, 3, 2, 0.5))
par(mgp = c(2.5, 1, 0))
par(cex.lab = 1.25)
par(mfrow=c(1,2))

x_logistic_1 <- logistic_sim(
  0.8,
  lambda = -1,
  K = 1
)

x_logistic_2 <- logistic_sim(
  0.01,
  lambda = 1,
  K = 1
)

plot(x_logistic_1,
     xlab = "step",
     ylab = "x",
     main = "lambda=-1, K=1",
     col = "dodgerblue",
     pch = 20,
     ylim = c(0,1)
)
abline(h = 0, col = "orange", lwd=2)

plot(x_logistic_2,
     xlab = "step",
     ylab = "x",
     main = "lambda=1, K=1",
     col = "dodgerblue",
     pch = 20,
     ylim = c(0,1)
)
abline(h = 1, col = "orange", lwd=2)
```
